\thetitle

### 2013-09-29 tfpie/response.txt

----------------------- REVIEW 1 ---------------------

I use the term "work-in-progress," because the authors have
side-stepped the methodology used to introduce types and the
transition from their tower of student languages to an industrial
language. Granted, this is likely something the authors are still
polishing. Nonetheless, it is important to provide a vision for how to
do these things and to outline what they have done highlighting both
succeses and failures. In this manner, interested readers may join
their efforts by reproducing the authors' methodology in their schools
using the languages developed by the authors. Let's face it,
introducing types is necessary to make the move to an
industrial-strength language and without providing a vision for it
many may balk. I suspect that the transition builds on the informal
contracts students are familiar with. Thus, adding this vision should
not be to difficult. Outlining this vision may obviate doubts readers
may have about making the transition to an industrial-strength
language. This is the primary reason that I am unable to put this
article in the strong accept column.

RESPONSE: We have added a paragraph in section 4.2 on this subject.

Page 2

Spell out an example of the details of numeric representation that
crop up quickly.

RESPONSE: We have added an example.

Implicit in the discussion is the assumption that reaching industrial
language competence by the end of the semester is a desirable
goal. The authors should address this head on. An obvious alternative
is emphasize and reinforce OO concepts and leave the industrial
language for latter (one semester?) in the curriculum. I am aware that
at NE students are expected to get a co-op/intership over the
summer. That, however, is a NE pecularity and is not a pedagogical
argument.

RESPONSE: We make explicit in section 2 the aspects of Northeastern
that inform our course.  It is not only co-op, but downstream courses,
that require Java.

Page 3

"sent by using the send keyword, followed by an object, a message
name, and some number of arguments"

        What are these "some number of arguments?"
        They are not part of the examples.

RESPONSE: These particular methods take no arguments, but the `dist`
method, on the same page, takes an argument.

Page 6

Spell out for the reader how students are explained the world
parameter is not needed and, therefore, removed from the methods in
the class definition.

RESPONSE: This is already covered in the discussion of OO
fundamentals.

In the class version, how does the runtime know which method to use?
Are the names of the handlers restricted to on-tick, on-key, etc.? Do
the methods have to appear in a certain order? Does the mechanism used
to make the association of events with the proper handler pose any
difficulty for the students? One nice feature of the functional
approach is that such associations are made explicitly and this allows
students to name their handlers whatever makes sense to them.

RESPONSE: We made this more explicit in the paper.

Do you mean the first interactive event-driven program?

RESPONSE: No, this is the first program we show.

The programs in Figure 2 do not animate a rocket. Please reformulate
the description of the programs in Figure 2.

RESPONSE: We have revised the program.

Show the reader how you use with students in class the state pattern
to animate a rocket that lands. There is plenty of room within the
20-page limit.

RESPONSE: We added an appendix with this program.

Page 11

Spell out the drawbacks that are alleviated and those that are not
addressed by DrJava and BlueJ. As a reader, I want to clearly see the
contrast of your approach with these.

Page 12

Can you quantify the recruitment/retention impact on female students?
Give the reader a sense of what you mean. It must be significant given
that you mention it!

Minor comments

Overall: Capitalize major words in section/subsection headers (not
just the first word).

Page 3
        "and message are sent" --> "and messages are sent"

Page 5
        "figure 1" --> "Figure 1"

Page 6

        "the structure of this data definition" --> "the structure of
        this class definition"

        "design of a game thus involves" --> "design of a game, thus,
        involves"

        "2htdp/universe" over the margin

        "figure 2" --> "Figure 2"

        "rockets, one" --> "rockets and one"

        "language levels", --> "language levels,"
Page 8
        "figure 3" --> "Figure 3"

----------------------- REVIEW 2 ---------------------
PAPER: 1
TITLE: From Principles to Practice with Class in the First Year
AUTHORS: Sam Tobin-Hochstadt and David Van Horn

OVERALL EVALUATION: 2 (accept)
REVIEWER'S CONFIDENCE: 4 (high)

----------- REVIEW -----------

I enjoyed reading this paper. It describes the approach to teaching
programming at Northeastern University, which begins by teaching
Racket in the first semester, followed by an introduction to
programming with classes in the second semester, initially by
extending Racket with classes, then moving on to a functional subset
of Java, and finally "traditional" Java.  The aim is to introduce
important programming concepts step by step, while still providing
students with enough practical knowledge to participate in an
internship. In particular, it is a nice observation that it is only a
small shift in focus to move from HDTP style to object oriented style
by adding a new language concept which the students already (almost)
understand.

The paper sets out the background and the constraints under which the
course has been designed, and does this well. It describes informally
the experience of both students and teachers on the course, and
suggests that the results have been positive. It is easy to read and
proceeds at a good pace, and I think it deserves to be published in
order to encourage more institutions to consider designing courses in
this way.

Some specific questions/comments/typos:

p1 "approache" => approach

p2 "a set that is demographically skewed" - in what way?  Also, is
this first paragraph written from your own experience of past versions
of the course?

p5 Would "figure 1", "figure 2" etc be better capitalised?

p5 I think this is the first time the names "BSL" and "Class" are
mentioned.  As far as I know the name BSL is taken from HTDP, and
there does seem to be a few places where it is assumed the reader is
familiar with the content of HTDP. Maybe it would be good to give a
brief summary of the HTDP approach and what these languages are
earlier.

p6 Interesting (and somewhat surprising to me) comment just before
section 3.4 that some invariants are easier to establish in the object
oriented style.

p7 Is this the first mention of "Intermediate Student" language?
Again, maybe in section two where the idea of language levels is
mentioned, maybe it would be good to go into a bit more detail for the
benefit of those who haven't read HTDP.

p9 Is it possible to include the Racket version alongside this
program? I found it interesting to compare them, and there does seem
to be enough space on the page for it.

p10 "functional OO programming" reads a bit strangely to me, since it
could mean "functional" in a non-technical sense. Maybe "a functional
subset of ..."?

p12 It's an interesting comment about retaining female students, but
it's not clear how the design of the course might be relevant to
this. Do you have any suggestions why? Could it just be a random
fluctuation?

p13 Some of the capitalisation in the references has gone ('java'
appears a few times for example)

----------------------- REVIEW 3 ---------------------
PAPER: 1
TITLE: From Principles to Practice with Class in the First Year
AUTHORS: Sam Tobin-Hochstadt and David Van Horn

OVERALL EVALUATION: 1 (weak accept)
REVIEWER'S CONFIDENCE: 4 (high)

----------- REVIEW -----------

In this paper the authors present their approach to smoothen the
transition during the first year programming course from semester 1:
FP to semester 2: OO programming. The key contribution of their work
is to demonstrate how a pure FP approach can be extended step by step
with OO concepts. In the end students acquire the same (and even
higher) level of proficiency of programming OO in Java as comparable
other courses.

The project's success relies on DrRacket, used in semester 1, which
can be extended with the desired features step by step. Halfway during
semester 2, when a sufficient 'critical mass' of OO concepts has been
introduced, the course and the students switch to Java and Eclipse.

The paper is well written. The paper is clearly in scope of TFPIE.

My main motivation for marking the paper as 'weak accept' is that in
my opinion it is not (yet) complete to deserve publication in
EPTCS. My main criticism plus advice for improvement is:

(1) in the introduction, page 1, the bullet-list enumerates
'significant transitions happen in concert'. In the paper we learn how
to overcome bullet 1, but not bullets 2, 3, 4, and 5. On page 8, first
paragraph: "In particular: objects, classes, fields and methods,
dynamic dispatch, inheritance, and overriding." are enumerated as
being essential concepts of object-oriented programming that students
have had experience with in the new set-up. The paper explains this
for the first half ('objects, classes, fields and methods') but not
for the second half ('dynamic dispatch, inheritance, and overriding').

Suggestion for improvement: extend section 3 with material how dynamic
dispatch, inheritance, and overriding is presented to students.

(2) Section 5.1 is too long to my taste: a lot of the criticism has
been discussed in the motivation of the introduction of this paper.

Suggestion for improvement: shorten section 5.1. The extra space can
be used to elaborate on issue (1) above. I don't think this section
requires sub sections, so all text can be organized into a single
section.

(3) In the introduction, it is said that 'an alternative approach to
the second semester [that] overcomes these problems and provides a
gradual introduction to OO programming'. However, acting as the
devil's advocate, I can argue that this is not true: halfway semester
2 there is still the abrupt switch to both another language, Java,
*and* another tool set, Eclipse. Stated bluntly, the switch is only
postponed instead of solved.

Suggestion for improvement: if issue (1) is resolved, it will become
clearer how the 'switch-gap' is made smaller. Section 4.1 enumerates
the differences of this gap, but does not explain how the students are
taught to jump over them. This should be added in this
section. Sections 4.2 and 4.3 are too brief in my opinion. Are the
topics in these sections taken immediately from 'How to Design
Classes'? If so, please state this. If not, please include material
that illustrates how these topics are taught and how this fits in the
philosophy of the chosen approach.

Minor remarks:
page 1:
"approache" --> "approach"

"numerous observers [9, 23]": only two references, of which [23] is
not scientific is not particularly 'numerous'. Please provide more
scientific references or remove the word 'numerous' (I prefer the
first option)

page 2:
"as follows: in section 2" --> "as follows: section 2"

page 4: the use of ";=> 3" and ";=> 4" in the same teletype font is
confusing because they are not part of the expression, but probably
intended to denote 'evaluates to'? Please use $math$ here.

page 5: "the BSL language and the Class language." these languages
have not been introduced yet, please do.

----------------------- REVIEW 4 ---------------------
PAPER: 1
TITLE: From Principles to Practice with Class in the First Year
AUTHORS: Sam Tobin-Hochstadt and David Van Horn

OVERALL EVALUATION: 2 (accept)
REVIEWER'S CONFIDENCE: 4 (high)

----------- REVIEW -----------

The authors describe a new approach to tackle difficulties in
transitioning from a functional semester one course to an
object-oriented semester two course. Students are reportedly confused
by the circumstantial differences (compilation model / environment /
syntax) which makes grasping the fundamental concepts being taught
harder to grasp. Instead of transitioning on day one of the second
semester to Java in Eclipse (or similar), the authors present an
approach that continues the progression of student languages to
include objects and other OO concepts before transitioning to a "real"
OO environment and language.

The paper is well-written, is well-positioned in the context of the
status quo and balances the pedagogical approach with the material
used. The idea of language levels is properly motivated and the
continuation of this approach beyond the first semester seems - after
reading the paper - the Obvious Choice. What is lacking, however, is a
motivation for the point at which the transition to Java is made. Why
aren't e.g. iteration, mutability & typing introduced in the similar
vein as objects, class-hierarchies, etc.? Was this a choice from
practicality, or is there a reason to introduce them post-transition?

A few specific comments and minor corrections:

 - (Sec.1) "common approache to" --> "common approach to"

 - (Sec.1) "interaction-oriented language" / "batch-oriented language"
   --> it seems the authors are refering to the environments/tooling
   rather than the language. If not, this terminology warrents
   clarification.

 - (Sec.1) "aspects of course remain" --> "aspects of *the* course
   remain"

 - (Sec.2) "while most students do some software development, there is
   almost no uniformity beyond that" & "in which they will use
   professional-grade languages and tools which will almost certainly
   be object-oriented" are, at least, paradoxical (if not
   contradictory). This conflict in the motivation should be addressed
   for the final version of the paper.

 - (Sec.3.1) "and message are sent" --> "and message*s* are sent"

 - (Sec.4.2) "Thanks to the the preparation" --> omit one "the"

- (Sec.6) "The class has also had notable success in the recruitment
  and retention of female students, as compared to the other versions
  of the second-semester course." --> I find this a rather hefty
  statement. It would - in my mind - at least warrent further
  research. If this consistently holds true, there are valuable
  insights to be gained about pedagogical gender-script in the field
  of CS. The statement by itself seems a bit hefty to be included
  without at least some concrete data and/or qualitative evaluation.

### 2013-09-29 tfpie/tfpie.tex
progress

\usepackage[numbers]{natbib}
\usepackage{xspace}

@@

\item from a language with numeric values corresponding to
  mathematical objects to a language with numeric values corresponding
  to common machine representations,\footnote{While this may seem like
    a minor point, details of numeric representation can crop up
    quickly in a Java-based course.}

@@

The key difference between these two programs is that the functional
program uses the {\tt 2htdp/universe} library, which provides a {\tt
  big-bang} form that consumes the initial state of the world and has
a declarative form of associating event-handler functions, while the
object-oriented program uses an alternative library developed for our
class: {\tt class/universe}.  It also provides a {\tt big-bang} form
but it consumes a single argument, the initial state of the world
represented as an object.  Event handlers are simply methods of this
object.

@@

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets, one for landed rockets.  While in the
functional approach it is easy to use the state-pattern for the
\emph{data} representing a rocket, it is more difficult to have states
of \emph{behavior}.  The typical solution adds conditionals to all of
the event handlers.  In the object-oriented approach, states of
behavior are just as natural as data.  It is therefore straightforward
to design programs with easy to observe invariants such as ``a landed
rocket never changes position.''  In the functional approach, even
such simple properties are more involved to establish.

@@

;; to-draw : World -> Image
(define (draw w)
  (circle w "solid" "red"))

@@

  ;; to-draw : -> Image
  (define (to-draw) 
    (circle (this . n) "solid" "red"))

@@

% \nocite{*}
\bibliographystyle{plainnat}
\bibliography{dvh-bibliography,local,sth-bibliography}
\end{document}

>>>>

\usepackage[numbers]{natbib}
\usepackage{xspace}
\usepackage{alltt}

@@

\item from a language with numeric values corresponding to
  mathematical objects to a language with numeric values corresponding
  to common machine representations,\footnote{While this may seem like
    a minor point, details of numeric representation can crop up
    quickly in a Java-based course---for example, $1/3$ cannot be
    represented by any Java numeric types.}

@@

The key difference between these two programs is that the functional
program uses the {\tt 2htdp/universe} library, which provides a {\tt
  big-bang} form that consumes the initial state of the world and has
a declarative form of associating event-handler functions, while the
object-oriented program uses an alternative library developed for our
class: {\tt class/universe}.  It also provides a {\tt big-bang} form
but it consumes a single argument, the initial state of the world
represented as an object.  Event handlers are simply methods of this
object; for example, clock ticks trigger the {\tt on-tick} method.

@@

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets, one for landed rockets (an example is given in
appendix~\ref{sec:appendix}).  While in the functional approach it
is easy to use the state-pattern for the \emph{data} representing a
rocket, it is more difficult to have states of \emph{behavior}.  The
typical solution adds conditionals to all of the event handlers.  In
the object-oriented approach, states of behavior are just as natural
as data.  It is therefore straightforward to design programs with easy
to observe invariants such as ``a landed rocket never changes
position.''  In the functional approach, even such simple properties
are more involved to establish.

@@

;; to-draw : World -> Image
(define (draw w)
  (place-image 
    (circle 10 "solid" "red")
    w 200 (empty-scene 400 400)))

@@

  ;; to-draw : -> Image
  (define (to-draw) 
    (place-image 
     (circle 10 "solid" "red")
     (this . w) 200 (empty-scene 400 400)))

@@

% \nocite{*}
\bibliographystyle{plainnat}
\bibliography{dvh-bibliography,local,sth-bibliography}

\newpage
\appendix
\section{Worlds and the State pattern}
\label{sec:appendix}

\begin{alltt}
#lang class/1
(require 2htdp/image class/universe)

;; A World is one of 
;; - (new upworld Number)
;; - (new downworld Number)
(define-class upworld
  (fields n)

  ;; to-draw : -> Image
  (define (to-draw) 
    (place-image 
     (circle 10 "solid" "red")
     (this . w) 200 (empty-scene 400 400)))

(define-class downworld
  (fields n)

  ;; on-tick : -> World
  (define (on-tick)
    (cond [(zero? (this . n))
           (new upworld (this . n))]
          [else
           (new world (sub1 (this . n)))]))

  ;; to-draw : -> Image
  (define (to-draw) 
    (place-image 
     (circle 10 "solid" "red")
     (this . w) 200 (empty-scene 400 400)))

(big-bang (new downworld 400))
\end{alltt}

\end{document}
